The aims of science and an open science controversy at USEPA

### 2025-05-09 paper/paper.qmd
enough content is done for ISEE; need to work on refs

format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: false
    number-sections: true
    link-citations: true
    pdf-engine: "lualatex"
    mainfont: "Times New Roman"
    fontsize: "12pt"
    geometry: 
      - "margin = 1.25in"
    linestretch: 2
  aft-pdf:
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    include-in-header:
        - text: |
            \usepackage{pdflscape}
    number-sections: true
    cite-method: citeproc

@@

execute:
  eval: true
  echo: false
  warning: true
  message: false
---

@@

While not as influential as the argument from inductive risk *[Douglas]*, the aims approach to values in science provides a distinct challenge to the value-free ideal. Proponents of the approach argue that science typically has not only epistemic aims but also practical or pragmatic ones, and that at any stage of inquiry decisions should be made so as to promote those aims. So, specifically, in the "core" of inquiry decisions should be made so as to promote pragmatic aims, and in this way non-epistemic values can play a legitimate role in the "core" of inquiry. *[cites]*  For example, *[Elliott and McKaughan]* argue that the pragmatic aim of quick and protective regulatory review supports using *[high-throughput]* methods and a potentially greater false positive rate in chemical safety testing. 
Note that this argument does not necessarily deny that science has epistemic aims, as may be done by approaches inspired by John Dewey *[Anderson; Brown]* or Donald Davidson *[Clough]*. This argument is also independent of questions about the definition of epistemic values or a sharp demarcation of values into "epistemic" vs. "social" *[Longino; Rooney; McMullin; Steel]*. Instead, the key assumption of this argument is that science does not have only epistemic aims.

@@

At least in the initial public notice, Strengthening Transparency was rationalized by an appeal to the open science movement and as a necessary response to the "replication crisis," a purportedly widespread crisis in which influential and high-profile scientific results often could not be replicated [@USEPAStrengtheningTransparencyRegulatory2018 18769-70].  @HicksOpenScienceReplication2023 argued that, at the time ST was proposed, there was limited evidence that the replication crisis extended beyond a handful of scientific fields, and specifically there was no evidence that the crisis covered environmental epidemiology.

@@

Text mining is an emerging method for empirical philosophy of science.  Broadly, text mining aims to identify patterns in the use of terms across a large collection of documents (called a *corpus*).  While initially developed by computer scientists, text mining methods have been widely adopted by both social scientists [@NelsonComputationalGroundedTheory2020; @StoltzMappingTextsComputational2024] and — under the umbrella of "digital humanities" — humanists [@UnderwoodDistantHorizonsDigital2019].

@@

We propose that public comments on high-profile governmental regulatory actions such as Strengthening Transparency can be apt and convenient corpora for understanding the publics involved in public scientific controversies.  In the United States, the Administrative Procedure Act requires federal government agencies to follow a "notice and comment" process before adopting new or modified regulations *[cite]*.  Agencies are required to publish a Notice of Proposed Rulemaking (NPR), including the draft text of the rule and background information to justify it, and then collect and respond to public comments before finalizing the rule.  Conveniently for our purposes, most comments are made freely and publicly available.  High-profile proposed rules can collect large numbers of comments, from a variety of different groups and perspectives on the issue at hand in the rule, making it apt for understanding how different groups might, for instance, deploy different views of the aims of science.

Most likely due to the mainstream news coverage, ST had an exceptionally high profile, with nearly one million comments submitted.^[In the text mining results below, the corpus contains "only" about 22,000 comments.  Regulations.gov distinguishes between "comments posted" and "comments received"; a single comment for the "posted" count can include multiple comments for the "received" count.  For example, the longest comments in the ST corpus comprise tens of thousands of signed electronic form letters.  This aggregation would count as one "posted" comment but tens of thousands of "received" comments.]  We use this substantial corpus to examine how supporters and opponents talk about science and health, finding that opponents talk about health and the environment about as often as they talk about science; while supporters talk exclusively about science, with a distinctive emphasis on a cynical or skeptical framing of the epistemic reliability of the science utilized by USEPA.

We argue that these differences between opponents and supporters correspond to Hicks' [-@HicksWhenVirtuesAre2022] distinction between broad and narrow views of science.  *[more]*

@sec-methods-overview gives a non-technical overview of our text mining methods; @sec-methods provides a more technical discussion, for readers who are familiar with text mining.  @sec-results reports the empirical results of the analysis. In @sec-discussion, we first argue that our empirical results can be fruitfully understood in terms of the two views of the aims of science.  We go on to suggest that, in the context of public discourse, deployment of philosophical ideas should be understood as somewhat opportunistic interpretive frames, rather than authentic expressions of beliefs or deep philosophical commitments.  Finally, we consider the role of advocacy organizations in shaping the content of public scientific controversies, and consider the applicability of a "participatory" model developed in misinformation studies to the case of ST.

@@

The US General Services Administration receives and publishes public comments on proposed rules using the website Regulations.gov (<https://www.regulations.gov>).  Commenters may submit comments by online form, email, or physical mail, and electronic formats allow commenters to include a number of attachments.  Attachments are typically detailed comments or to include journal articles, reports, etc., as supporting evidence.

@@

Initial, undirected exploration of the corpus in 2020-2021 by [A1] indicated that (a) there was a qualitatively clear division between supporters and opponents of ST, (b) commenters offered substantive reasons for their position on ST, (c) these reasons often connected science to public health, but (d) opposing comments were much, much more prevalent than supportive comments.  The point (a) meant that manual qualitative coding could be used to identify supporting and opposing comments; combined with points (b) and (c), this meant that the corpus was likely apt for contrasting the arguments given on either side of the ST controversy, and especially the way they connected (or not) epistemic and pragmatic aims.  But the third point meant that a simple random sample would include very few supporters.  @sec-sampling details the sampling procedure that we used for the manual coding step.

@@

```{r}
#| eval: true
#| label: "tbl-n-docs"
#| tbl-cap: "Document counts, by coding method and opposition/support"
#| tbl-subcap:
#|   - "Whole corpus"
#|   - "Comments with science/health bigrams"
#| layout-cols: 2
read_rds(here(out_dir, '09_n_docs.Rds'))
# read_rds(here(out_dir, '09_n_docs_bigrams.Rds'))
```

@@

There are also notable differences in the overall tone of the adjectives in supporters' and opponents' top bigrams.  Supporters seem to express skepticism or cynicism about the quality of scientific research: science is described as secret, secretive, phony, false, fake, shady, or deceptive.  In the context of these skeptical assessments, seemingly-positive valence bigrams — such as real, true, verifiable, and sound science — might be used as privatives, that is, expressing the idea that actually-existing science is *not* real, true, verifiable, or sound.  A skeptical or cynical assessment of science would provide support for ST insofar as open data would make science less secretive and phony and more verifiable and true, etc.  @HicksOpenScienceReplication2023 argued that there is no empirical evidence of a replication crisis in environmental epidemiology [though see @BagiletEstimatingSmallEffects *[cite as well in intro]*], and that ST's open data requirements would not mitigate widespread irreplicability even if it were a problem in that field. @HicksDevelopingMeasuresPublic2024 report preliminary evidence that cynicism is a better predictor of generalized trust in science than the value-free ideal.

@@

Deployment of the broad view might also explain why opponents said relatively little about business and the economy.  On Hicks' theory of value, protecting human health and the environment are aims of environmental public health, but the impacts of regulation on the economy are still accidents, and thus irrelevant to the standards of good scientific research.  @HicksValuesDisclosuresTrust2022 provide survey evidence that, in the context of chemical regulation, members of the public — including political conservatives — put more trust in a scientist who discloses public health values than one who does not disclose values or discloses economic values.

@@

Applied to the ST controversy, we suggest that both liberals and conservatives have both narrow and broad views of the aims of science among their repertoire of frames for interpreting public scientific controversies. *[narrow view is depoliticizing; broad view is politicizing?]*

Reading the public comments as expressions of interpretive frames rather than beliefs can also help us understand what the "authors" of form letter comments were doing when they submitted a comment without modifying or even carefully reading the suggested text. For the low-stakes action of submitting a comment, details of argument and wording were less important than the general sentiment of the text — its vibes. So a few seconds to skim the suggested text would be sufficient for most commentators.

## The role of advocacy organizations

*[Starbird: participatory disinformation]*

>>>>

format:
  html:
    toc: true
    number-sections: true
  aft-pdf:
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    include-in-header:
        - text: |
            \usepackage{pdflscape}
            \usepackage{framed}
    number-sections: true
    cite-method: citeproc

@@

execute:
  eval: true
  echo: false
  warning: false
  message: false
---

@@

While not as influential as the argument from inductive risk *[Douglas, Havstad]*, the aims approach to values in science provides a distinct challenge to the value-free ideal. Proponents of the approach argue that science typically has not only epistemic aims but also practical or pragmatic ones, and that at any stage of inquiry decisions should be made so as to promote those aims. So, specifically, in the "core" of inquiry decisions should be made so as to promote pragmatic aims, and in this way non-epistemic values can play a legitimate role in the "core" of inquiry. *[cites]*  For example, *[Elliott and McKaughan]* argue that the pragmatic aim of quick and protective regulatory review in chemical safety testing supports using rapid screening techniques, which are much faster than other methods but will likely have a higher false positive rate.

Note that this argument does not necessarily deny that science has epistemic aims, as may be done by approaches inspired by John Dewey *[Anderson; Brown]* or Donald Davidson *[Clough]*. This argument is also independent of questions about the definition of epistemic values or a sharp demarcation of values into "epistemic" vs. "social" *[Longino; Rooney; McMullin; Steel]*. Instead, the key assumption of this argument is that science does not have *only* epistemic aims.

@@

At least in the initial public notice, proponents of Strengthening Transparency appealed to the open science movement and, specifically, open science as a necessary response to the "replication crisis," a purportedly widespread crisis in which influential and high-profile scientific results often could not be replicated [@USEPAStrengtheningTransparencyRegulatory2018 18769-70].  @HicksOpenScienceReplication2023 argued that, at the time ST was proposed, there was limited evidence that the replication crisis extended beyond a handful of scientific fields, and specifically there was no evidence that the crisis covered environmental epidemiology.

@@

Text mining has emerged as a method for empirical philosophy of science over the past decade.  Broadly, text mining aims to identify patterns in the use of terms across a large collection of documents (called a *corpus*).  While initially developed by computer scientists, text mining methods have been widely adopted by both social scientists [@NelsonComputationalGroundedTheory2020; @StoltzMappingTextsComputational2024] and — under the umbrella of "digital humanities" — humanists [@UnderwoodDistantHorizonsDigital2019].

@@

We propose that public comments on high-profile governmental regulatory actions such as Strengthening Transparency can be apt and convenient corpora for understanding the publics involved in public scientific controversies.  In the United States, the Administrative Procedure Act requires federal government agencies to follow a "notice and comment" process before adopting new or modified regulations *[cite]*.  Agencies are required to publish a Notice of Proposed Rulemaking (NPR), including the draft text of the rule and background information to justify it, and then collect and respond to public comments before finalizing the rule.  Conveniently, most comments are made freely and publicly available on the website `regulations.gov`, which also includes a free API (application programming interface) to facilitate bulk downloads ("scraping").  High-profile proposed rules can collect large numbers of comments, from a variety of different groups and perspectives on the issue at hand in the rule, making it apt for understanding how different groups might, for instance, deploy different views of the aims of science.

## Text mining ST: Primary findings and roadmap of the paper

Presumably due to the mainstream news coverage, ST had an exceptionally high profile, with nearly one million comments submitted and available on `regulations.gov`.^[In the text mining results below, the corpus contains "only" about 22,000 comments.  Regulations.gov distinguishes between "comments posted" and "comments received"; a single comment for the "posted" count can include multiple comments for the "received" count.  For example, the longest comments in the ST corpus comprise tens of thousands of signed electronic form letters.  This aggregation would count as one "posted" comment but tens of thousands of "received" comments.]  We use this substantial corpus to examine how supporters and opponents talk about science and health, finding that opponents talk about health and the environment about as often as they talk about science; while supporters talk exclusively about science, with a distinctive emphasis on a cynical or skeptical framing of the epistemic reliability of the science utilized by USEPA.

We argue that these differences between opponents and supporters correspond to Hicks' [-@HicksWhenVirtuesAre2022] distinction between broad and narrow views of science.

@sec-methods-overview gives a non-technical overview of our text mining methods; @sec-methods provides a more technical discussion, for readers who are familiar with text mining.  @sec-results reports the empirical results of the analysis. In @sec-discussion, we first argue that our empirical results can be fruitfully understood in terms of the two views of the aims of science.  We go on to suggest that, in the context of public discourse, deployment of philosophical ideas — such as views of the aims of science — might best be understood as somewhat opportunistic interpretive frames, rather than expressions of beliefs.

@@

The US General Services Administration receives and publishes public comments on proposed rules using the website `regulations.gov` (<https://www.regulations.gov>).  Commenters may submit comments by online form, email, or physical mail, and electronic formats allow commenters to include a number of attachments.  Attachments are typically longer comments or journal articles, reports, etc., as supporting evidence.

@@

Initial, undirected exploration of the corpus in 2020-2021 by [A1] indicated that (a) there was a qualitatively clear division between supporters and opponents of ST, (b) commenters offered substantive reasons for their position on ST, (c) these reasons often connected science to public health, but (d) opposing comments were much, much more prevalent than supportive comments.  Example comments are given in figures @fig-opposed and @fig-support.

::: {#fig-opposed}
\begin{framed}
\raggedright

Im writing to speak out AGAINST your proposed regulations to assault the EPAs use of science in the middle of a global public health crisis.

Peer-reviewed studies assessing the effects of pollution exposure on humans, based on actual personal health data which by law must be kept confidential are among the best science available to the Agency. The misguided proposal would direct EPA to ignore these data unless those confidential personal health data were made publicly available.

This is both unnecessary and impractical. As a retired scientist, I urge you to abandon this reckless plan, for the sake of our childrens health. To do otherwise is unconscionable.
\end{framed}

Example of an opposing comment on Strengthening Transparency. Original available at <https://www.regulations.gov/comment/EPA-HQ-OA-2018-0259-19410>. 
:::
::: {#fig-support}
\begin{framed}
\raggedright
I sincerely hope that you can distance the EPA agency from politically correct science. The only way to do this is to adopt the following methodology.

The scientific method demands that researchers' work be fully transparent.

Data must be made available for examination.

Theories must be open to challenge.

Disproven theories must be altered or set aside.

Barriers must be removed that would prevent valid conclusions from rising to the top.

I for one am tired of hearing about how bad technologies are for the planet, without any scientific proof that things today are different. Don't tell me that glarciers are melting and the ocean is rising a foot every 10 years. Remove yourselves from such statements and give facts about such a change: volume of water, how will it be measured, what could be causing it, what can be done that is both environmentally and economically sound.
\end{framed}

Example of a supporting comment on Strengthening Transparency. Original available at <https://www.regulations.gov/comment/EPA-HQ-OA-2018-0259-12305>.
:::

Point (a) meant that manual qualitative coding could be used to identify supporting and opposing comments; combined with points (b) and (c), this meant that the corpus was likely apt for contrasting the arguments given on either side of the ST controversy, and especially the way they connected (or not) epistemic and pragmatic aims.  But (d) meant that a simple random sample would include very few supporters, which would be suboptimal for using machine learning to automatically code the entire corpus.  @sec-sampling details the sampling procedure that we used for the manual coding step.

@@

```{r}
#| eval: true
#| label: "tbl-n-docs"
#| tbl-cap: "Document counts, by coding method and opposition/support"
#| tbl-subcap:
#|   - "Whole corpus"
#|   - "Comments with science/health bigrams"
#| layout-ncol: 1
read_rds(here(out_dir, '09_n_docs.Rds'))
read_rds(here(out_dir, '09_n_docs_bigrams.Rds'))
```

@@

*[move most of this to discussion]*
There are also notable differences in the overall tone of the adjectives in supporters' and opponents' top bigrams.  Supporters seem to express skepticism or cynicism about the quality of scientific research: science is described as secret, secretive, phony, false, fake, shady, or deceptive.  In the context of these skeptical assessments, seemingly-positive valence bigrams — such as real, true, verifiable, and sound science — might be used as privatives, that is, expressing the idea that actually-existing science is *not* real, true, verifiable, or sound.  A skeptical or cynical assessment of science would provide support for ST insofar as open data would make science less secretive and phony and more verifiable and true, etc.  @HicksOpenScienceReplication2023 argued that there is no empirical evidence of a replication crisis in environmental epidemiology [though see @BagiletEstimatingSmallEffects *[cite as well in intro]*], and that ST's open data requirements would not mitigate widespread irreplicability even if it were a problem in that field. @HicksDevelopingMeasuresPublic2024 report preliminary evidence that cynicism is a better predictor of generalized trust in science than the value-free ideal.

@@

Deployment of the broad view might also explain why opponents said relatively little about business and the economy.  On Hicks' analysis protecting human health and the environment are aims of environmental public health, but the impacts of regulation on the economy are still accidents, and thus irrelevant to the standards of good scientific research.  @HicksValuesDisclosuresTrust2022 provide survey evidence that, in the context of chemical regulation, members of the public — including political conservatives — put more trust in a scientist who discloses public health values than one who does not disclose values or discloses economic values.

@@

Applied to the ST controversy, we suggest that both liberals and conservatives have both narrow and broad views of the aims of science among their repertoire of frames for interpreting public scientific controversies. But the frames were evoked differently in this particular controversy by being connected to two very different frames on the general trustworthiness of EPA/regulatory science *[McCright and Dunlap]*

While the degree of conservative distrust in science is often exaggerated *[cites]*, there is a large difference in partisan attitudes towards USEPA: @CerdaAmericansFeelFavorably2024 reports that only 32% of Republican/Lean Republican survey respondents have a favorable attitude towards USEPA, compared to 73% of Democratic/Lean Democratic respondents. Republican distrust of USEPA appears in the bigram analysis *[xref]*, in the *way* supporters talk about science, using adjectives such as *[list]*. These characterizations of science portray it as suffering from an inappropriate influence of non-epistemic values — specifically, politicization *[cite]*. The narrow view and the value-free ideal justify outrage at this situation, and by extension policies such as ST that are (believed to) mitigate these inappropriate influences. We further suggest that liberal utilization of the narrow view, in the context of scientific integrity plans, is also a response to politicization.

More generally, the narrow view might be evoked or deployed when science is (perceived as) being influenced by *values one disagrees with*. This can be understood as what we might call a *depoliticization rhetorical strategy*, in two senses: first, as a call to mitigate the influence of politics; but second as an appeal to premises (epistemic values) that are taken to be uncontroversial and external to politics. In the second sense, depoliticization is a version of proxy politics [@HicksScientificControversiesProxy2017] or what Sarewitz calls the "scientization" of politics [@SarewitzHowScienceMakes2004], where political actors present political concerns and arguments as apolitical and purely scientific.

*[broad view]*

Finally, reading the public comments as expressions of interpretive frames rather than beliefs can also help us understand what the "authors" of form letter comments were doing when they submitted a comment without modifying or even carefully reading the suggested text. For the low-stakes action of submitting a comment, details of argument and wording were less important than the general sentiment of the text — its vibes. So a few seconds to skim the suggested text would be sufficient for most commentators.
